Predicting trends in the quality of state-of-the-art neural networks without access to training or testing data

### 2020-02-16 submission/methods.tex
full pass, fixed lots of typos, etc.; semifinal, nearly done

Assume we are given several pretrained DNNs, e.g., as part of an architecture series.
The models have been trained and evaluated on labeled data $\{d_{i},y_{i}\}\in\mathcal{D}$, using standard techniques.  
The pretained pytroch model files are publicly-available, and the test accuracies have been reported online.  
In this study, we do not have access to this data, and we have not trained any of the models ourselves,
nor have we re-evaluated the test accuracies.
We expect that most well-trained, production-quality models will employ one or more forms of regularization, such as Batch Normalization (BN), Dropout, etc., and many will also contain additional structure such as Skip Connections, etc. 
Here, we will ignore these details, and will focus only on the pretrained layer weight matrices $\mathbf{W}_{l}$.

@@

The \emph{Weighted Alpha metric} is an average of $\alpha_l$ over all layers $l \in \{1,\ldots,l\}$, weighted by the size, or scale, or each matrix,
%(and it approximates the average log $\alpha$-Shatten Norm metric),
\begin{equation}
\hat{\alpha} = \dfrac{1}{L}\sum_l \alpha_l\log\lambda_{max,l}\approx\langle\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}\rangle    ,
\end{equation}
where $L$ is the total nuumber of layer weight matrices.
The Weighted Alpha metric was introduced previously~\cite{MM20_SDM}, where it was shown to correlate well with trends in reported test accuracies of pretrained DNNs, albeit on a limited set of models.

>>>>

Assume we are given several pretrained DNNs, e.g., as part of an architecture series.
The models have been trained and evaluated on labeled data $\{d_{i},y_{i}\}\in\mathcal{D}$, using standard techniques.  
The pretrained pytroch model files are publicly-available, and the test accuracies have been reported online.  
In this study, we do not have access to this data, and we have not trained any of the models ourselves,
nor have we re-evaluated the test accuracies.
We expect that most well-trained, production-quality models will employ one or more forms of regularization, such as Batch Normalization (BN), Dropout, etc., and many will also contain additional structure such as Skip Connections, etc. 
Here, we will ignore these details, and will focus only on the pretrained layer weight matrices $\mathbf{W}_{l}$.

@@

The \emph{Weighted Alpha metric} is an average of $\alpha_l$ over all layers $l \in \{1,\ldots,l\}$, weighted by the size, or scale, or each matrix,
%(and it approximates the average log $\alpha$-Shatten Norm metric),
\begin{equation}
\hat{\alpha} = \dfrac{1}{L}\sum_l \alpha_l\log\lambda_{max,l}\approx\langle\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}\rangle    ,
\end{equation}
where $L$ is the total number of layer weight matrices.
The Weighted Alpha metric was introduced previously~\cite{MM20_SDM}, where it was shown to correlate well with trends in reported test accuracies of pretrained DNNs, albeit on a limited set of models.

### 2020-02-16 submission/nlp_models.tex
full pass, fixed lots of typos, etc.; semifinal, nearly done

In this section, we examine empirical quality metrics described in Section~\ref{sxn:methods} for several NLP model architectures.
%
%In particular, 
Within the past two years, nearly 100 open source, pretrained NLP DNNs based on the revolutionary Transformer architecture have emerged.
These include variants of BERT, Transformer-XML, GPT, etc.
%
The Transformer architectures consist of blocks of so-called Attention layers, containing two large, Feed Forward (Linear) weight matrics~\cite{Attn2017}. 
In contrast to smaller pre-Activation maps arising in Cond2D layers, Attention matrices are significantly larger.
In general, we have found that they have larger PL exponents $\alpha$.
Based on HT-SR Theory, 
%(in particular, the interpretation of values of $\alpha \sim 2$ as modeling systems with good correations over many size scales~\cite{BouchaudPotters03, SornetteBook}), 
this suggests that these models fail to capture successfully many of the correlations in the data (relative to their size) and thus are substantially \emph{under}-trained.
%
More generally, compared to the CV models of Section~\ref{sxn:cv},
modern NLP models have larger weight matrices and display different spectral properties.
%, both in terms of their architecture (e.g., wheter there are convolutions) as well as how they are used (e.g., whether one uses prediction error or perplexity).
Thus, they provide a very different test for our empirical quality metrics.

While norm-based metrics perform reasonably well on well-trained NLP models, they often behave anomalously on poorly-trained models.
Indeed, for such ``bad'' models, weight matrics may display rank collapse, decreased Frobeinus mass, or unsually small Spectral norms.
(This may be misinterpreted as ``smaller is better.'')
In contrast, PL-based metrics, including the Log $\alpha$-Norm metric ($\log\Vert\mathbf{W}\Vert_{\alpha}^{\alpha}$) and the Weighted Alpha metric ($\hat\alpha =\alpha\log\lambda_{max} $) display consistant behavior, even on poorly trained models.
Indeed, we can use these metrics to help indentify when archiectures need repair and when more and/or better data are needed.

@@

The OpenAI GPT and GPT2 models provide us with the opportunity to analyze two effects: training the same model with different data set sizes; and increasing the sizes of both the data set and the architectures simultaneously.
These models have the remarkable ability to generate fake text that appears to the human to be real, and they have generated significant media attention because of the potential for their misuse.
For this reason, the original GPT model released by OpenAI was trained on a deficient data set, rendering the model interesting but not fully functional.  
Later, OpenAI released a much improved model, GPT2-small, which has the same architecture and number of layers as GPT, but which has been trained on a larger and better data set (and with other changes), making it remarkably good at generating (near) human-quality fake text.  
%
By comparing the poorly-trained (i.e., ``bad'') GPT to the well-trained (i.e., ``good'') GPT2-small, we can indentify empirical indicators for when a model has in fact been poorly-trained and thus may perform poorly when deployed.
By comparing GPT2-medium to GPT2-large to GPT2-xl, we can examine the effect of increasing data set and model size simultaneously,
an example of what we call a series of ``good-better-best'' models.

The GPT models we analyze are deployed with the popular HuggingFace PyTorch library~\cite{huggingface}.
GPT has 12 layers, with 4 Multi-head Attention Blocks, giving $48$ layer Weight Matrices, $\mathbf{W}$.
Each Block has 2 components, the Self Attention (attn) and the Projection (proj) matrics.  
The self-attention  matrices are larger, of dimension ($2304\times 768$) or ($3072\times 768$).
The projection layer concatenates the self-attention results into a vector (of dimension $768$).
This gives $50$ large matrices.
%
Because GPT and GPT2 are trained on different data sets, the initial Embedding matrices differ in shape.
GPT has an initial Token and Positional Embedding layers, of dimension $(40478\times 768)$ and $(512\times 768)$, respectively, whereas GPT2 has input Embeddings of shape $(50257\times 768)$ and $(1024\times 768)$, respectively. 
%
The OpenAI GPT2 (English) models are: GPT2-small, GPT2-medium, GPT2-large, and GPT2-xl, having $12$, $24$, $36$, and $48$ layers, respectively, with increasingly larger weight matrices.
%The model card for GPT2 is published on github.\footnote{\url{https://github.com/openai/gpt-2/blob/master/model_card.md}}.

\begin{table}[t]
\small
\begin{center}
\begin{tabular}{|p{1in}|c|c|c|c|c|}
%KDD% \begin{tabular}{|p{0.67in}|c|c|c|c|c|}
\hline
 Series  & \#   & $\langle\log\Vert\mathbf{W}\Vert_{F}\rangle$ & $\langle\log\Vert\mathbf{W}\Vert_{\infty}\rangle$ & $\hat{\alpha}$ & $\langle\log\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}\rangle$ \\
\hline
GPT & 49 & 1.64  & 1.72 & 7.01 & 7.28 \\
GPT2-small & 49 & 2.04  & 2.54& 9.62 & 9.87 \\
\hline
GPT2-medium & 98 & 2.08 & 2.58& 9.74 & 10.01 \\
GPT2-large & 146 & 1.85 & 1.99& 7.67 & 7.94 \\
GPT2-xl & 194 & 1.86 & 1.92 & 7.17 & 7.51 \\
\hline
\end{tabular}
\end{center}
\caption{Average value for the average Log Norm and Weighted Alpha metrics for pretrainnd OpenAI GPT and GPT2 models. Column \# refers to number of layers treated.  Note that the averages do not include the first embedding layer(s) because they are not (implicitly) normalized.  }
\label{table:nlp}
\end{table}

@@

Going beyond average values, Figure~\ref{fig:GPT-alpha-hist} shows the histogram (empirical density), for all layers, of $\alpha$ for GPT and GPT2-small.  
These two histograms are very different.
The older deficient GPT has numerous unsually large $\alpha$ exponents---meaning they are not really well-described by a PL fit.
Indeed, we expect that a poorly-trained model will lack good (i.e., small $\alpha$) PL behavior in many/most layers.
On the other hand, as expected, the newer improved GPT2-small model has, on average, smaller $\alpha$ values than the older GPT, with all $\alpha\le6$ and with smaller mean/median $\alpha$.
It also has far fewer unusually-large outlying $\alpha$ values than GPT.
From this (and other results not shown), we see that $\alpha$ provides a good quality metric for comparing these two models, the ``bad'' GPT versus the ``good'' GPT2-small.
%
This should be contrasted with the behavior displayed by the Frobenius norm (not shown) and the Spectral~norm.

@@

We also examine in Figure~\ref{fig:gpt-alpha-layers} the PL exponent $\alpha$ and Log Spectral Norm versus layer id, for GPT and GPT2-small.
Let's start with Figure~\ref{fig:gpt-alpha-layer}, which plots $\alpha$ versus the depth (i.e., layer id) for each model.
The deficient GPT model displays two trends in $\alpha$, one stable with $\alpha\sim 4$, and one increasing with layer id, with $\alpha$ reaching as high as $12$.
In contrast, the well-trained GPT2-small model shows consistant and stable patterns, again with one stable $\alpha\sim 3.5$ (and below the GPT trend), and the other only slightly trending up, with $\alpha\le 6$. 
The scale-invariant $\alpha$ metric lets us identify potentially poorly-trained models.
These results show that the Correlation Flow differs significantly between GPT and GPT2-small (with the better GPT2-small looking more like the better ResNet-1K from Figure~\ref{fig:resnet-alpha-layer}).

@@

We see that the average $\alpha$ decreases with increasing model size, although the differences are less noticible between the differing good-better-best GTP2 models than between the good-versus-bad GPT and GPT2-small models.
Unlike GPT, however, the layer Log Alpha Norms behave more as expected for GPT2 layers, with the larger models consistently having smaller norms. 
Similarly, the Log Spectral Norm also decreases on average with the larger models (not shown).  
As expected, the norm metrics can indeed distinguish among good-better-best models among a series well-trained models.

@@

    \subfigure[PL exponent ($\alpha$)]{
        %KDD% \includegraphics[width=4.0cm]{img/GPT2*_fnl_alpha_hist.png}
        \includegraphics[width=5.5cm]{img/GPT2*_fnl_alpha_hist.png}
        \label{fig:gpt2-alpha-hist}
    }
    %KDD% 
    \quad
    \subfigure[Log Alpha Norm]{
        %KDD% \includegraphics[width=4.0cm]{img/GPT2*_fnl_logpnorm_hist.png}
        \includegraphics[width=5.5cm]{img/GPT2*_fnl_logpnorm_hist.png}
        \label{fig:gpt2-pnorm-hist}
    }
    \caption{Histogram of PL exponents ($\alpha$)
             %Log Spectral norm ($\log\Vert\mathbf{W}\Vert_{\infty}$), 
              and Log Alpha Norm ($\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}$) for weight matrices from models of different sizes in the GPT2 architecture series.  (Plots omit the first 2 (embedding) layers, because they are normalized differently giving anamolously large values.)
             }
    \label{fig:gpt2-histograms}
\end{figure}

>>>>

In this section, we examine empirical quality metrics described in Section~\ref{sxn:methods} for several NLP model architectures.
%
%In particular, 
Within the past two years, nearly 100 open source, pretrained NLP DNNs based on the revolutionary Transformer architecture have emerged.
These include variants of BERT, Transformer-XML, GPT, etc.
%
The Transformer architectures consist of blocks of so-called Attention layers, containing two large, Feed Forward (Linear) weight matrices~\cite{Attn2017}. 
In contrast to smaller pre-Activation maps arising in Cond2D layers, Attention matrices are significantly larger.
In general, we have found that they have larger PL exponents $\alpha$.
Based on HT-SR Theory
(in particular, the interpretation of values of $\alpha \sim 2$ as modeling systems with good correlations over many size scales~\cite{BouchaudPotters03, SornetteBook}), 
this suggests that these models fail to capture successfully many of the correlations in the data (relative to their size) and thus are substantially \emph{under}-trained.
%
More generally, compared to the CV models of Section~\ref{sxn:cv},
modern NLP models have larger weight matrices and display different spectral properties.
%, both in terms of their architecture (e.g., whether there are convolutions) as well as how they are used (e.g., whether one uses prediction error or perplexity).
Thus, they provide a very different test for our empirical quality metrics.

While norm-based metrics perform reasonably well on well-trained NLP models, they often behave anomalously on poorly-trained models.
Indeed, for such ``bad'' models, weight matrices may display rank collapse, decreased Frobenius mass, or unusually small Spectral norms.
(This may be misinterpreted as ``smaller is better.'')
In contrast, PL-based metrics, including the Log $\alpha$-Norm metric ($\log\Vert\mathbf{W}\Vert_{\alpha}^{\alpha}$) and the Weighted Alpha metric ($\hat\alpha =\alpha\log\lambda_{max} $) display consistent behavior, even on poorly trained models.
Indeed, we can use these metrics to help identify when architectures need repair and when more and/or better data are needed.

@@

The OpenAI GPT and GPT2 models provide us with the opportunity to analyze two effects: training the same model with different data set sizes; and increasing the sizes of both the data set and the architectures simultaneously.
These models have the remarkable ability to generate fake text that appears to the human to be real, and they have generated significant media attention because of the potential for their misuse.
For this reason, the original GPT model released by OpenAI was trained on a deficient data set, rendering the model interesting but not fully functional.  
Later, OpenAI released a much improved model, GPT2-small, which has the same architecture and number of layers as GPT, but which has been trained on a larger and better data set (and with other changes), making it remarkably good at generating (near) human-quality fake text.  
%
By comparing the poorly-trained (i.e., ``bad'') GPT to the well-trained (i.e., ``good'') GPT2-small, we can identify empirical indicators for when a model has in fact been poorly-trained and thus may perform poorly when deployed.
By comparing GPT2-medium to GPT2-large to GPT2-xl, we can examine the effect of increasing data set and model size simultaneously,
an example of what we call a series of ``good-better-best'' models.

The GPT models we analyze are deployed with the popular HuggingFace PyTorch library~\cite{huggingface}.
GPT has 12 layers, with 4 Multi-head Attention Blocks, giving $48$ layer Weight Matrices, $\mathbf{W}$.
Each Block has 2 components, the Self Attention (attn) and the Projection (proj) matrices.  
The self-attention  matrices are larger, of dimension ($2304\times 768$) or ($3072\times 768$).
The projection layer concatenates the self-attention results into a vector (of dimension $768$).
This gives $50$ large matrices.
%
Because GPT and GPT2 are trained on different data sets, the initial Embedding matrices differ in shape.
GPT has an initial Token and Positional Embedding layers, of dimension $(40478\times 768)$ and $(512\times 768)$, respectively, whereas GPT2 has input Embeddings of shape $(50257\times 768)$ and $(1024\times 768)$, respectively. 
%
The OpenAI GPT2 (English) models are: GPT2-small, GPT2-medium, GPT2-large, and GPT2-xl, having $12$, $24$, $36$, and $48$ layers, respectively, with increasingly larger weight matrices.
%The model card for GPT2 is published on github.\footnote{\url{https://github.com/openai/gpt-2/blob/master/model_card.md}}.

\begin{table}[t]
\small
\begin{center}
\begin{tabular}{|p{1in}|c|c|c|c|c|}
%KDD% \begin{tabular}{|p{0.67in}|c|c|c|c|c|}
\hline
 Series  & \#   & $\langle\log\Vert\mathbf{W}\Vert_{F}\rangle$ & $\langle\log\Vert\mathbf{W}\Vert_{\infty}\rangle$ & $\hat{\alpha}$ & $\langle\log\Vert\mathbf{X}\Vert^{\alpha}_{\alpha}\rangle$ \\
\hline
GPT & 49 & 1.64  & 1.72 & 7.01 & 7.28 \\
GPT2-small & 49 & 2.04  & 2.54& 9.62 & 9.87 \\
\hline
GPT2-medium & 98 & 2.08 & 2.58& 9.74 & 10.01 \\
GPT2-large & 146 & 1.85 & 1.99& 7.67 & 7.94 \\
GPT2-xl & 194 & 1.86 & 1.92 & 7.17 & 7.51 \\
\hline
\end{tabular}
\end{center}
\caption{Average value for the average Log Norm and Weighted Alpha metrics for pretrained OpenAI GPT and GPT2 models. Column \# refers to number of layers treated.  Note that the averages do not include the first embedding layer(s) because they are not (implicitly) normalized.  }
\label{table:nlp}
\end{table}

@@

Going beyond average values, Figure~\ref{fig:GPT-alpha-hist} shows the histogram (empirical density), for all layers, of $\alpha$ for GPT and GPT2-small.  
These two histograms are very different.
The older deficient GPT has numerous unusually large $\alpha$ exponents---meaning they are not really well-described by a PL fit.
Indeed, we expect that a poorly-trained model will lack good (i.e., small $\alpha$) PL behavior in many/most layers.
On the other hand, as expected, the newer improved GPT2-small model has, on average, smaller $\alpha$ values than the older GPT, with all $\alpha\le6$ and with smaller mean/median $\alpha$.
It also has far fewer unusually-large outlying $\alpha$ values than GPT.
From this (and other results not shown), we see that $\alpha$ provides a good quality metric for comparing these two models, the ``bad'' GPT versus the ``good'' GPT2-small.
%
This should be contrasted with the behavior displayed by the Frobenius norm (not shown) and the Spectral~norm.

@@

We also examine in Figure~\ref{fig:gpt-alpha-layers} the PL exponent $\alpha$ and Log Spectral Norm versus layer id, for GPT and GPT2-small.
Let's start with Figure~\ref{fig:gpt-alpha-layer}, which plots $\alpha$ versus the depth (i.e., layer id) for each model.
The deficient GPT model displays two trends in $\alpha$, one stable with $\alpha\sim 4$, and one increasing with layer id, with $\alpha$ reaching as high as $12$.
In contrast, the well-trained GPT2-small model shows consistent and stable patterns, again with one stable $\alpha\sim 3.5$ (and below the GPT trend), and the other only slightly trending up, with $\alpha\le 6$. 
The scale-invariant $\alpha$ metric lets us identify potentially poorly-trained models.
These results show that the Correlation Flow differs significantly between GPT and GPT2-small (with the better GPT2-small looking more like the better ResNet-1K from Figure~\ref{fig:resnet-alpha-layer}).

@@

We see that the average $\alpha$ decreases with increasing model size, although the differences are less noticeable between the differing good-better-best GTP2 models than between the good-versus-bad GPT and GPT2-small models.
Unlike GPT, however, the layer Log Alpha Norms behave more as expected for GPT2 layers, with the larger models consistently having smaller norms. 
Similarly, the Log Spectral Norm also decreases on average with the larger models (not shown).  
As expected, the norm metrics can indeed distinguish among good-better-best models among a series well-trained models.

@@

    \subfigure[PL exponent ($\alpha$)]{
        %KDD% \includegraphics[width=4.0cm]{img/GPT2*_fnl_alpha_hist.png}
        \includegraphics[width=5.5cm]{img/GPT2*_fnl_alpha_hist.png}
        \label{fig:gpt2-alpha-hist}
    }
    %KDD% 
    \quad
    \subfigure[Log Alpha Norm]{
        %KDD% \includegraphics[width=4.0cm]{img/GPT2*_fnl_logpnorm_hist.png}
        \includegraphics[width=5.5cm]{img/GPT2*_fnl_logpnorm_hist.png}
        \label{fig:gpt2-pnorm-hist}
    }
    \caption{Histogram of PL exponents ($\alpha$)
             %Log Spectral norm ($\log\Vert\mathbf{W}\Vert_{\infty}$), 
              and Log Alpha Norm ($\log\Vert\mathbf{X}\Vert_{\alpha}^{\alpha}$) for weight matrices from models of different sizes in the GPT2 architecture series.  (Plots omit the first 2 (embedding) layers, because they are normalized differently giving anomalously large values.)
             }
    \label{fig:gpt2-histograms}
\end{figure}
